---
title: "Labelr - Special Topics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Labelr - Special Topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  error = TRUE,
  comment = "### >"
)
```

## Overview
While the Introductory vignette gives an overview of core labelr functions, 
here, we offer an ad hoc dive into a range of miscellaneous special topics and 
additional functionalities. Let's go. 

## Larger Data Frames
labelr is not intended for "large" data.frames, which is a fuzzy concept. To
give a sense of what labelr **can** handle, let's see it in action with the 
NYC Flights 2013 data set: a moderate-not-big data.frame of ~340K rows.

Let's load labelr and the nycflights13 package.
```{r}
opening_ding <- Sys.time() # to time labelr

library(nycflights13)
```

We'll assign the data.frame to one we call df.
```{r}
df <- flights

nrow(df)
```
We'll add a "frame label," which describes the data.frame overall.
```{r}
df <- add_frame_lab(df, frame.lab = "On-time data for all flights that
                    departed NYC (i.e. JFK, LGA or EWR) in 2013.")
```
Let's see what this did.
```{r}
attr(df, "frame.lab") # check for attribute

get_frame_lab(df) # return frame.lab alongside data.frame name as a data.frame

get_frame_lab(df)$frame.lab
```
Now, let's assign variable NAME labels.

```{r}
names_labs_vec <- c(
  "year" = "Year of departure",
  "month" = "Month of departure",
  "year" = "Day of departure",
  "dep_time" = "Actual departure time (format HHMM or HMM), local tz",
  "arr_time" = "Actual arrival time (format HHMM or HMM), local tz",
  "sched_dep_time" = "Scheduled departure times (format HHMM or HMM)",
  "sched_arr_time" = "Scheduled arrival time (format HHMM or HMM)",
  "dep_delay" = "Departure delays, in minutes",
  "arr_delay" = "Arrival delays, in minutes",
  "carrier" = "Two letter airline carrier abbreviation",
  "flight" = "Flight number",
  "tailnum" = "Plane tail number",
  "origin" = "Flight origin airport code",
  "dest" = "Flight destination airport code",
  "air_time" = "Minutes spent in the air",
  "distance" = "Miles between airports",
  "hour" = "Hour of scheduled departure time",
  "minute" = "Minutes component of scheduled departure time",
  "time_hour" = "Scheduled date and hour of the flight as a POSIXct date"
)

df <- add_name_labs(df, name.labs = names_labs_vec)

get_name_labs(df) # show that they've been added
```

Let's add variable VALUE labels for variable "carrier." Helpfully, this ships 
with the nycflights13 package itself.

```{r}
airlines <- nycflights13::airlines

head(airlines)
```

The carrier field of airlines matches the carrier column of df (formerly, flights)
```{r}
ny_val <- airlines$carrier
```

The name field of airlines gives us the full airline names.
```{r}
ny_lab <- airlines$name
```

df (flights) also has an integer month variable. We will "hand-jam" month value 
labels

```{r}
ny_month_vals <- c(1:12) # values
ny_month_labs <- c(
  "JAN", "FEB", "MAR", "APR", "MAY", "JUN",
  "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"
) # labels
```

Let's add these value labels. First, we'll demo `add_val1()`, then 
`add_val_labs()`, then `add_quant_labs()`.

```{r}
df <- add_val1(df,
  var = carrier, vals = ny_val,
  labs = ny_lab,
  max.unique.vals = 20
)
```

```{r}
df <- add_val_labs(df,
  vars = "month",
  vals = ny_month_vals,
  labs = ny_month_labs,
  max.unique.vals = 20
)
```

```{r}
df <- add_quant_labs(df, "dep_time", qtiles = 5)
```

Let's see where this leaves us.

```{r}
get_val_labs(df)
```

We can use `head()` to get a baseline look at select rows and variables
```{r}
head(df[c("origin", "dep_time", "dest", "year", "month", "carrier")])
```

Now, let's do the same for a version we modified with `use_val_labs()`. Note that
this cannot be "undone" (except for the usual clunky way of re-running our script
up to this point and not doing this!).

```{r}
df_swapd <- use_val_labs(df)

head(df_swapd[c("origin", "dep_time", "dest", "year", "month", "carrier")])
```

Instead of replacing values (which we can't undo), it might be safer to simply
add "value-labels-on" character variables to the data.frame. This adds nearly
1M new cells, but let's throw caution to the wind with `add_lab_cols()`.

```{r}
df_plus <- add_lab_cols(df, vars = c("carrier", "month", "dep_time"))

head(df_plus[c(
  "origin", "dest", "year",
  "month", "month_lab",
  "dep_time", "dep_time_lab",
  "carrier", "carrier_lab"
)])
```

We can use `flab()` to filter df based on month and carrier, even when value
labels are "invisible" (i.e., existing only as attributes() meta-data.

```{r}
# labels are not visible (they exist only as attributes() meta-data)
head(df[c("carrier", "arr_delay")])

# we still can use them to filter (note: we're filtering on "JetBlue Airways",
# ...NOT its obscure code "B6")
df_fl <- flab(df, carrier == "JetBlue Airways" & arr_delay > 20)

# here's what's returned when we filtered on "JetBlue Airways" using flab()
head(df_fl[c("carrier", "arr_delay")])

# double-check that this is JetBlue
head(use_val_labs(df_fl)[c("carrier", "arr_delay")])
```

How long did this entire NYC Flights session take (results will vary)?
```{r}
the_buzzer <- Sys.time()
the_buzzer - opening_ding
```

## NA and "Irregular" Values
labelr is no fan of NA values or other "irregular" values, which are defined as
infinite values, not-a-number values, and character values that look like them
(e.g., "NAN", "INF", "inf", "Na"). 

When value-labeling a column / variable, such values are automatically given the 
catch-all label "NA" (which will be converted to an actual NA in any columns 
created by `add_lab_cols()` or `use_val_labs()`). You do not need (and should 
not try) to specify this yourself, and you should not try to over-ride labelr on 
this. If you want to use labelr AND you present with these sorts of values, your 
options are to accept the default "NA" label or convert these sorts of values to 
something else before labeling. 

With that said, let's see how labelr handles this, with an assist from our old
friend mtcars (packaged with R's base distribution).

First, let's assign mtcars to a new data.frame object that we will besmirch.
```{r}
mtbad <- mtcars
```

Let's get on with the besmirching.
```{r}
mtbad[1, 1:11] <- NA
rownames(mtbad)[1] <- "Missing Car"
mtbad[2, "am"] <- Inf
mtbad[3, "gear"] <- -Inf
mtbad[5, "carb"] <- NaN
mtbad[2, "mpg"] <- Inf
mtbad[3, "mpg"] <- NaN

# add a character variable, for demonstration purposes
# if it makes you feel better, you can pretend these are Consumer Reports or
# ...JD Power ratings or something
set.seed(9202) # for reproducibility
mtbad$grade <- sample(c("A", "B", "C"), nrow(mtbad), replace = TRUE)
mtbad[4, "grade"] <- NA
mtbad[5, "grade"] <- "NA"
mtbad[6, "grade"] <- "Inf"

# see where this leaves us
head(mtbad)

sapply(mtbad, class)
```
Now, let's add value labels to this unruly data.frame. 

```{r}
mtlabs <- mtbad |>
  add_val1(grade,
    vals = c("A", "B", "C"),
    labs = c("Gold", "Silver", "Bronze")
  ) |>
  add_val1(am,
    vals = c(0, 1),
    labs = c("auto", "stick")
  ) |>
  add_val1(carb,
    vals = c(1, 2, 3, 4, 6, 8), # not the most inspired use of labels
    labs = c(
      "1c", "2c", "3c",
      "4c", "6c", "8c"
    )
  ) |>
  add_val1(gear,
    vals = 3:5, # again, not the most compelling use case
    labs = c(
      "3-speed",
      "4-speed",
      "5-speed"
    )
  ) |>
  add_quant1(mpg, qtiles = 4) # add quartile-based value labels
```

```{r}
get_val_labs(mtlabs, "am") # NA values were detected and dealt with
```

Let's streamline the data.frame with `sselect()` to make it more manageable.

```{r}
mtless <- sselect(mtlabs, mpg, cyl, am, gear, carb, grade) # safely select

head(mtless, 5) # note that the irregular values are still here
```

Notice how all irregular values are coerced to NA when we substitute labels for 
values with `use_val_labs()`.

```{r}
head(use_val_labs(mtless), 5) # but they all go to NA if we `use_val_labs`
```

Now, let's try an `add_lab_cols()` view.

```{r}
mtlabs_plus <- add_lab_cols(mtlabs, c("mpg", "am")) # this creates and adds "am_lab" col
mtlabs_plus <- sselect(mtlabs_plus, mpg, mpg_lab, am, am_lab) # let's select down to these two

head(mtlabs_plus) # here's where we landed
```

What if we had tried to explicitly label the NA values and/or irregular values 
themselves? We would have failed. 

```{r}
# Trying to Label an Irregular Value (-Inf)
mtbad <- add_val1(
  data = mtcars,
  var = gear,
  vals = -Inf,
  labs = c("neg.inf")
)

# Trying to Label an Irregular Value (NA)
mtbad <- add_val_labs(
  data = mtbad,
  vars = "grade",
  vals = NA,
  labs = c("miss")
)

# Trying to Label an Irregular Value (NaN)
mtbad <- add_val_labs(
  data = mtbad,
  vars = "carb",
  vals = NaN,
  labs = c("nan-v")
)

# labelr also treats "character variants" of irregular values as irregular values.
mtbad <- add_val1(
  data = mtbad,
  var = carb,
  vals = "NAN",
  labs = c("nan-v")
)
```

Again, labelr handles NA and irregular values and resists our efforts to take 
such matters into our own hands.

## Factors and Value Labels
R's concept of a factor variable shares some affinities with the concept of a
value-labeled variable and can be viewed as one approach to value labeling. 
However, factors can manifest idiosyncratic and surprising behaviors depending 
on the function to which you're trying to apply them. They are character-like, 
but they are not character values. They are built on top of integers, but they 
won't submit to all of the operations that integers do. They do some very handy 
things in certain model-fitting applications, but their behavior "under the hood" 
can be counter-intuitive or opaque. Simply put, they are their own thing.

So, while factors have their purposes, it would be nice to associate value 
labels with the distinct values of data.frame variables in a manner that 
preserves the integrity and transparency of the underlying values (factors tend 
to be a bit opaque about this) and that allows you to view or use the labels in 
flexible ways.

And if you wanted to work with a factor, it would be nice if you could add value 
labels to it without it ceasing to exist and behave as a factor. 

### Adding Labels to a Factor
With that said, let's see if we can have our label-factor cake and eat it, too, 
using the iris data.frame that comes pre-packaged with R. 

```{r}
unique(iris$Species)

sapply(iris, class) # nothing up our sleeve -- "Species" is a factor
```

Let's add value labels to "Species" and assign the result to a new data.frame 
that we'll call irlab. For our value labels, we'll use "se", "ve", and "vi", 
which are not adding much new information, but they will help to illustrate what 
we can do with labelr and a factor variable.

```{r}
irlab <- add_val_labs(iris,
  vars = "Species",
  vals = c("setosa", "versicolor", "virginica"),
  labs = c("se", "ve", "vi")
)

# this also would've worked
# irlab_dos <- add_val1(iris, Species,
#   vals = c("setosa", "versicolor", "virginica"),
#   labs = c("se", "ve", "vi")
# )
```

Note that we could have just as (or even more) easily used `add_val1()`, which 
works for a single variable at a time and allows us to avoid quoting our column 
name, if that matters to us. In contrast, `add_val_labs()` requires us to put 
our variable name(s) in quotes, but it also gives us the option to apply a 
common value-label scheme to several variables at once (e.g., Likert-style 
survey responses). We'll see an example of this type of use case in action in a 
little bit. 

For now, though, let's prove that the iris and irlab data.frames are 
functionally identical.

First, note that irlab looks and acts just like iris in the usual ways that 
matter

```{r}
summary(iris)

summary(irlab)

head(iris, 4)

head(irlab, 4)

lm(Sepal.Length ~ Sepal.Width + Species, data = iris)

lm(Sepal.Length ~ Sepal.Width + Species, data = irlab) # values are same
```

Note also that irlab's "Species" is still a factor, just like its iris 
counterpart/parent.

```{r}
sapply(irlab, class)

levels(irlab$Species)
```

But irlab's "Species" has value labels!

```{r}
get_val_labs(irlab, "Species")
```

And they work.

```{r}
head(use_val_labs(irlab))
ir_v <- flab(irlab, Species == "vi")
head(ir_v, 5)
```

Our take-aways so far? Factors can be value-labeled while staying factors, and 
we can use the labels to do labelr-y things with those factors. We can have both.

We may want to go further and add the labeled variable alongside the factor version.

```{r}
irlab_aug <- add_lab_cols(irlab, vars = "Species")
```

This gives us a new variable called "Species_lab". Let's get select rows of the 
resulting data.frame, since we want to see all the different species.

```{r}
set.seed(231)
sample_rows <- sample(seq_len(nrow(irlab)), 10, replace = FALSE)

irlab_aug[sample_rows, ]

sapply(irlab_aug, class)

with(irlab_aug, table(Species, Species_lab))
```

Caution: Replacing the entire data.frame using `use_val_labs()` WILL coerce 
factors to character, since the value labels are character values, not 
recognized factor levels

```{r}
ir_char <- use_val_labs(irlab) # we assign this to a new data.frame
sapply(ir_char, class)

head(ir_char, 3)

class(ir_char$Species) # it's character
```

Of course, even then, we could explicitly coerce the labels to be factors if we 
wanted

```{r}
ir_fact <- use_val_labs(irlab)

ir_fact$Species <- factor(ir_char$Species,
  levels = c("se", "ve", "vi"),
  labels = c("se", "ve", "vi")
)
head(ir_fact, 3)

class(ir_fact$Species) # it's a factor

levels(ir_fact$Species) # it's a factor
```

We've recovered.

```{r}
with(ir_fact, tapply(Sepal.Width, Species, mean))
with(irlab, tapply(Sepal.Width, Species, mean))
with(iris, tapply(Sepal.Width, Species, mean))
```

### Ordered factors
Value labels work with ordered factors, too. Let's make a fictional ordered 
factor that we add to ir_ord. We can pretend that this is some sort of judge's 
overall quality rating, if that helps.

```{r}
ir_ord <- iris

set.seed(293)
qrating <- c("AAA", "AA", "A", "BBB", "AA", "BBB", "A")

ir_ord$qrat <- sample(qrating, 150, replace = TRUE)

ir_ord$qrat <- factor(ir_ord$qrat,
  ordered = TRUE,
  levels = c("AAA", "AA", "A", "BBB")
)
```

Where do we stand with this factor?

```{r}
levels(ir_ord$qrat)

class(ir_ord$qrat)
```

Now, let's add value labels to it.

```{r}
ir_ord <- add_val_labs(ir_ord,
  vars = "qrat",
  vals = c("AAA", "AA", "A", "BBB"),
  labs = c(
    "unimpeachable",
    "excellent",
    "very good",
    "meh"
  )
)
```

Let's add a separate column with those labels as a distinct (character) variable 
unto itself, existing in addition to (not replacing) "qrat".

```{r}
ir_ord <- add_lab_cols(ir_ord, vars = "qrat")

head(ir_ord, 10)

with(ir_ord, table(qrat_lab, qrat))

class(ir_ord$qrat)

levels(ir_ord$qrat)

class(ir_ord$qrat_lab)

get_val_labs(ir_ord, "qrat") # labs are still there for qrat

get_val_labs(ir_ord, "qrat_lab") # no labs here; this is just a character var
```

It appears that you really can have it all, where "it all" is defined as 
"factors and labels."

## Value-Labeling Many Variables at Once
Functions for adding value labels (e.g., `add_val_labs`, `add_quant_labs` and 
`add_m1_lab`) will do partial matching if the partial argument is set to TRUE. 
Let's use labelr's `make_likert_data()` function to generate some fake Likert 
scale-style survey data to demonstrate this more fully. 

```{r}
set.seed(272) # for reproducibility
dflik <- make_likert_data(scale = 1:7) # another labelr function
head(dflik)
```
We'll put the values we wish to label and the labels we wish to use in 
stand-alone vectors, which we will supply to `add_val_labs` in a moment.

```{r}
vals2label <- 1:7
labs2use <- c(
  "VSD",
  "SD",
  "D",
  "N",
  "A",
  "SA",
  "VSA"
)
```

Now, let's associate/apply the value labels to ALL vars with "x" in their name 
and also to var "y3." Note: partial = TRUE.

```{r}
dflik <- add_val_labs(
  data = dflik, vars = c("x", "y3"), ###  note the vars args
  vals = vals2label,
  labs = labs2use,
  partial = TRUE # applying to all cols with "x" or "y3" substring in names
)
```

Let's compare dflik with value labels present but "off" to labels "on."

First, present but "off."

```{r}
head(dflik)
```

Now, let's "turn on" (use) these value labels.
```{r}
lik1 <- uvl(dflik) # assign to new object, since we can't "undo"
head(lik1) # we could have skipped previous call by using labelr::headl(dflik)
```

Yea, verily: All variables with "x" in their name (and "y3") got the labels! 

Suppose we want to drop these value labels for a select few, but not all, of 
these variables. `drop_val_labs` can get the job done.

```{r}
dfdrop <- drop_val_labs(dflik,
  c("x2", "y3"),
  partial = FALSE
)
```

Most of our previously labeled columns remain so; but not "x2" and "y3."
```{r}
get_val_labs(dfdrop, c("x2", "y3"))
```

Compare to values for variable "x1" (we did not drop value labels from this one)
```{r}
get_val_labs(dfdrop, "x1")
```

Just like we did with `add_val_labs()`, we also can use a single command to drop 
value labels from all variables with "x" in their variable names.
```{r}
dfxgone <- drop_val_labs(dflik,
  c("x"),
  partial = TRUE # note
)
```

"y3" still has value labels, but now **all** "x" var value labels are gone.
```{r}
get_val_labs(dfxgone)
```

### Tabulating Frequencies with `tabl()`
Finally, let's get to know labelr's `tabl()` function, which supports count 
or proportion tabulations with labels turned "on" or "off" and offers some other 
functionalities. 

With `tabl()`, tables can be generated... 

...in terms of values
```{r}
head(tabl(df), 20) # labs.on = FALSE is default
```
...or in terms of labels
```{r}
head(tabl(df, labs.on = TRUE), 20) # labs.on = TRUE is not the default
```
...in proportions
```{r}
head(tabl(df, labs.on = TRUE, prop.digits = 3), 20)
```
...cross-tab style
```{r}
head(tabl(df, labs.on = TRUE, wide.col = "gender"), 20)
```
...with non-value-labeled data.frames
```{r}
tabl(iris, "Species") # explicit vars arg with one-var ("Species")

tabl(mtcars, zero.rm = TRUE) # vars arg null
```

## Conclusion
This is the suitably abrupt ending to our choppy, ad hoc overview of some 
additional labelr capabilities and special topics. Thanks for reading.

