---
title: "Labelr - An Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Labelr - An Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  error = TRUE,
  comment = "### >"
)
```

## A Whirlwind Tour
labelr supports creation and use of multiple types of labels for data.frames and 
their columns. This is an ad hoc introduction to core and ancillary labelr 
functionalities and uses cases. 

## Types of Labels
labelr supports the following kinds of labels:
  
1. **Frame labels** - Each data.frame may be given a single "frame label", which 
    can be used describe the data set's key general features or characteristics 
    (e.g., source, date produced or published, high-level contents). 
  
2. **Name labels** - Each data.frame column (variable) may be given exactly one 
    name label, which is an extended variable name or brief description of the 
    variable. Name labels equivalent to what Stata and SAS call "variable 
    labels."

3. **Value labels** - Specific values of a data.frame column (variable) can be 
    labeled as well. The package supports three (3) kinds of value labels. 
    * *One-to-one labels* - The canonical value-labeling use case entails mapping 
      distinct values of a variable to distinct labels in a one-to-one fashion, 
      so that each value label uniquely identifies a substantive value. For 
      instance, an administrative data set might assign the integers 1-7 to 
      seven distinct racial/ethnic groups, and value labels would be critical in 
      mapping those numbers to socially substantive racial/ethnic category 
      concepts (e.g., Which number corresponds to the category "Asian 
      American?"). 
    
    * *Many-to-one labels* - In an alternative use case, value labels may serve 
      to distill or "bucket" distinct variable values in a way that deliberately 
      "throws away" information for purposes of simplification. For example, one 
      may wish to give the single label "Agree" to the responses "Very Strongly 
      Agree," "Strongly Agree," and "Agree." Or one may wish to differentiate 
      self-identified "White" respondents from "People of Color," applying the 
      latter value label to all categories other than "White."
    
    * *Numerical range labels* - Finally, one may wish to carve a numerical 
      variable into qualitative bins, such as dichotomizing a variable or 
      dividing it into quantiles. Numerical range labels support one-to-many 
      assignment of a single value label to a range of numerical 
      values for a given variable.

## Core Use Cases and Capabilities
More specifically, labelr functions support the following actions:

1. Assigning variable value labels, name labels, and a frame label to data.frames 
   and modifying those labels thereafter.
   
2. Generating and accessing simple look-up table-style data.frames to inform or
   remind you about a data.frame's frame labels, its columns' name labels, or 
   the value labels that correspond to its unique values.
   
3. Swapping out variable (column) names for variable name labels and back again.

4. Replacing variables' actual values with their corresponding value labels.

5. Augmenting a data.frame by adding columns of variable value labels that can 
exist alongside the original columns (variables) from which they were derived.

6. Engaging in `base::subset()`-like row-filtering, using value labels to guide 
the filtering but returning a subsetted data.frame in terms of the original 
variable values.

7. Tabulating value frequencies that can be expressed in terms of raw values or
value labels -- again, without explicitly modifying or converting the raw 
data.frame values.

8. Preserving and restoring a data.frame's labels in the event that 
   some unsupported R operation destroys them.
   
9. Applying a single value-labeling scheme to many variables at once (for 
   example, assigning the same set of Likert-scale labels to all variables that 
   share a common variable name character substring).

## Disclaimer Regarding Base R Data Frames
Note: To minimize dependencies and reduce unexpected behaviors, many labelr 
functions **will coerce augmented/non-standard data.frames (e.g., tibbles, 
data.tables) to base R ("vanilla") data.frames.** If you work with non-standard 
data.frames such as these, the suggested workflow is to affix and use labelr 
labels **before** transforming the base R data.frame to a non-standard 
data.frame, if at all. While some augmented data.frames and their functions may 
"play well" with labelr-style labels and functions, this is not guaranteed. 
Experiment as desired and at your own discretion and risk.

## Adding and Looking up Frame, Name, and Value Labels
We'll start our exploration of core labelr functions with a fake "demographic" 
data.frame. First, though, let's load the package labelr.

### Load the Package
```{r setup}
# install.packages("labelr") #CRAN version
# install.packages("devtools") # Step 1 to get GitHub version
# devtools::install_github("rhartmano/labelr") #Step 2 to get GitHub version
library(labelr)
```

### Make Toy Demographic Data.Frame
We'll use `make_demo_data()` (included with labelr) to create the fictional 
data set.

```{r}
set.seed(555) # for reproducibility
df <- make_demo_data(n = 1000) # you can specify the number of fictional obs.

# make a backup for later comparison
df_copy <- df
```

### Add a Variable "FRAME label" Using `add_frame_lab()` 
We'll start our labeling session by providing a fittingly fictional high-level 
description of this fictional data set. labelr calls this a FRAME label.

```{r}
df <- add_frame_lab(df, frame.lab = "Demographic and reaction time test score
                    records collected by Royal Statistical Agency of
                    Fictionaslavica. Data fictionally collected in the year
                    1987. As published in A. Smithee (1988). Some Fictional Data
                    for Your Amusement. Mad Magazine, 10(1), 1-24.")


get_frame_lab(df)
```

### Add Variable NAME Labels Using `add_name_labs()` 
Now, let's add (some fairly trivial) variable NAME labels

```{r}
df <- add_name_labs(df, name.labs = c(
  "age" = "Age in years",
  "raceth" = "Racial/ethnic identity group category",
  "gender" = "Gender identity category",
  "edu" = "Highest education level attained",
  "x1" = "Space Invaders reaction time test scores",
  "x2" = "Galaga reaction time test scores"
))
```

Even if we do nothing else with these name labels, we can access or manipulate a
simple lookup table as needed.

```{r}
get_name_labs(df)
```

### Add VALUE labels Using `add_val_labs()`
Now, let's do some VALUE labeling. First, let's use `add_val_labs()` to add 
one-to-one value labels for the variable "raceth".  

```{r}
df <- add_val_labs(df, # data.frame with to-be-value-labeled column
  vars = "raceth", # quoted variable name of to-be-labeled col
  vals = c(1:7), # label values 1 through 7, inclusive
  labs = c(
    "White", "Black", "Hispanic", # ordered labels for vals 1-7
    "Asian", "AIAN", "Multi", "Other"
  ),
  max.unique.vals = 10 # max number of unique values permitted
)
```

### Add Value Labels Using `add_val1()`
Now let's add value labels for the variable "gender." Function `add_val1` is a 
variant of `add_val_labs` that allows you to supply the variable name unquoted, 
provided you are value-labeling only one variable. (It's not evident from the 
above, but `add_val_labs` supports labeling multiple variables at once).

```{r}
df <- add_val1(
  data = df,
  var = gender, # contrast this var argument to the vars argument demo'd above
  vals = c(0, 1, 2), # the values to be labeled
  labs = c("Male", "Female", "Other"), # the labels, applied in order to the vals
  max.unique.vals = 10
)
```

Once again, we can create a lookup table, this time for our labels-to-values 
mappings. Because we used `add_val_labs()` and `add_val`(), each unique value of 
our value-labeled variables will (must) have one unique label (one-to-one 
mapping), and any unique values that were not explicitly assigned a label were 
given one automatically (the value itself, coerced to character as needed).

```{r}
get_val_labs(df)
```

### Add NUMERICAL RANGE Labels Using `add_quant_labs()`
Traditionally, value labels are intended for categorical variables, such as 
binary, nominal, or (integer) ordinal variables with limited numbers of distinct 
categories. Further, as just noted, value labels that are added using 
`add_val_labs` (or `add_val1`) are constrained to map one-to-one to distinct 
values: No two distinct values could share a value label or vice versa.

If you wish to relax these constraints and apply a label to a range of values of 
a numeric variable, such as labeling each value according to the quintile or 
decile to which it belongs, you can use `add_quant_labs()` (or `add_quant1`) to 
do so.

Here, we will use `add_quant_labs` with the partial argument set to TRUE to 
apply quintile range labels to **all variables** of df that have an "x" in their 
names (i.e., vars "x1" and "x2"). We'll demonstrate this capability further at 
the end of this vignette.

```{r}
df_temp <- add_quant_labs(
  data = df,
  vars = "x",
  qtiles = 5,
  partial = TRUE
)

get_val_labs(df_temp)
```

**Be careful** with setting partial to TRUE like this: If your data set featured 
a column called "sex" or that featured the string "tax" or the suffix "max" in 
its name, `add_quant_labs()` would attempt to apply the value labeling scheme to 
that column as well! 

We can use the same function to assign arbitrary, user-specified range labels. 
Here, we assign numerical range labels based on an arbitrary cutpoint that 
differentiate values of "x1" and "x2" that are at or below 100 from values that 
are at or below 150 (but greater than 100).

```{r}
df_temp <- add_quant_labs(
  data = df_temp,
  vars = "x",
  vals = c(100, 150),
  partial = TRUE
)

get_val_labs(df_temp)
```

Having demonstrated the basic functionality on our df_temp copy of df, let's 
ignore that data.frame and return our focus to df. We'll use `add_quant1` to 
apply decile range labeling to the variable "x1" only. Note that `add_quant1` is 
like `add_quant_labs`, but accepts only a single variable, whose name can be 
supplied without quotes. The opposite trade-off holds for `add_quant_labs`: The 
relationship between these two functions mirrors the  relationship between 
`add_val_labs` and `add_val1`. 

```{r}
df <- add_quant1(df, # data.frame
  x1, # variable to value-label
  qtiles = 5
) # number quintiles to define numerical range labels
```

We'll preserve the "x1" range labels going forward, keeping "x2" unlabeled.

### Add MANY-TO-ONE VALUE Labels Using `add_m1_lab()`
If you wish to apply a single label to multiple distinct values that are not 
necessarily part of a numerical range, this can be done through successive calls 
to `add_m1_lab()` Here, the "m1" is shorthand for "many to one," as in "many 
values get the same one value label." 

Note that each call to `add_m1_lab()` applies a single value label, so,  
multiple calls are needed to apply multiple labels. Here, we illustrate this 
workflow, applying the label "Some College+" to values 3, 4, or 5 of the 
variable "edu", then applying other distinct labels to values 1 and 2, 
respectively.

```{r}
df <- add_m1_lab(df, "edu", vals = c(3:5), lab = "Some College+")
df <- add_m1_lab(df, "edu", vals = 1, lab = "Not HS Grad")
df <- add_m1_lab(df, "edu", vals = 2, lab = "HSG, No College")

get_val_labs(df)
```

As with the other value-adding functions, there is a variant of `add_m1_lab` 
that allows you to value-label a single variable whose name is unquoted. It is 
`add1m1()`.

### Where Do We Stand?
All of this is nice, but have we really accomplished anything? A casual view of 
the data.frame raises some doubts:
  
  ```{r}
head(df_copy, 3) # our pre-labeling copy of the data.frame

head(df, 3) # our latest, post-labeling version of same data.frame
```

These two data.frames still look identical. 

Rest assured, labeling has introduced some unobtrusive but important features 
for us to use. We'll put them to work in a moment. But first let's back them up 
in case we lose them. 

## Preserving and Restoring Labels
Lose them, you say? labelr labels are data.frame attributes, and certain Base R
functions (like some forms of subsetting) are known to destroy data.frame 
attributes. For this reason, once you're done labeling your data.frame, it's 
wise to create an in-session backup of your labels by assigning them to an 
extra, stand-alone object. You can do this with `get_all_lab_atts()`, which will 
return all labels (frame, name, and value) as a list object that you can 
subsequently (re-) attach to a data.frame, as needed.

```{r}
labs.df <- get_all_lab_atts(df)
```

Now, let's remove our labels explicitly, simulating what certain R functions do 
naturally.

```{r}
df <- strip_labs(df) # remove our labels
get_all_lab_atts(df) # show that they're gone
```

Now, let's restore them, using the labs.df list object we just created.

```{r}
df <- add_lab_atts(df, labs.df)

get_all_lab_atts(df)
```

We're back(ed up)!
  
In addition to this hack, labelr provides label-preserving variants of common 
data management functions, including `sfilter()`, `sselect()`, `ssubset()`, 
`srename()`, `ssort()`, and others. Here, the "s" prefix is for "safely," as in, 
"your labels will be safely retained"). Other popular packages (e.g., "dplyr") 
may also preserve label attributes, but an advantage of the labelr functions 
like `sselect()` is that they will update the label attributes of affected 
columns. For example, if your use of `sselect()` or `sdrop` removes a column 
from the returned data.frame, any labels associated with that column will be 
removed from that data.frame's attributes, as well. 

## "Using" Value Labels
Now that our data.frame is labeled (and our labels backed up), let's demonstrate 
some ways that we can use them.

### Show First, Last, or Random Rows with Value Labels Overlaid
Base R includes the `head()` and `tail()` functions, which allow you to show the
first n or last n rows of a data.frame. In addition, the "car" package offers a 
similar function called `some()`, which allows you to show a random n rows of a 
data.frame.

labelr provides versions of these functions that will display value labels in 
place of values, without actually altering the values in the underlying 
data.frame. Let's demonstrate each of the three standard functions, followed by 
its labelr counterpart. Note that the unconventional rownames (e.g., "T-1," 
"N-2") are provided as an aid to help you visually locate a literal row that may 
appear across calls.

```{r}
head(df, 5) # Base R function utils::head()

headl(df, 5) # labelr function headl() (note the "l")

tail(df, 5) # Base R function utils::tail()

taill(df, 5) # labelr function taill() (note the extra "l")

set.seed(293)
car::some(df, 5) # car package function car::some()

set.seed(293)
somel(df, 5) # labelr function somel() (note the "l")
```

Note that `some()` and `somel()` both return random rows, but they will not 
necessarily return the same random rows, even with the same random number seed.

### Swap out Values for Labels with `use_val_labs()` and `uvl()`
We can generalize this overlaying (aka "turning on" aka "swapping in") of value 
labels to the entire data.frame. For example, we might do this temporarily, to 
visualize the labels in place of values.

```{r}
use_val_labs(df)[1:20, ] # headl() is just a more compact shortcut for this
```

Or we can wrap a call to this function around our data.frame and pass the result 
to other functions. Here is an illustration that passes a `use_val_labs()` 
-wrapped data.frame to the `qsu()`function of the collapse package. To save 
typing, we'll use `uvl()`, a more compact alias for `use_val_labs()`. 

First we show the unwrapped call to `collapse::qsu()`, followed by an otherwise 
identical call that wraps the data.frame in `uvl()`. Focus your eyes on the 
leftmost column of the console outputs of the respective calls (i.e., the 
rownames of the object generated by `qsu::collapse()`).

```{r}
# `collapse::qsu()`
# with labels "off" (i.e., using regular values of "raceth" as by var)
(by_demog_val <- collapse::qsu(df, cols = c("x2"), by = ~raceth))

# with labels "on" (i.e., using labels, thanks to `uvl()`)
(by_demog_lab <- collapse::qsu(uvl(df), cols = c("x2"), by = ~raceth))
```

This second call would achieve the same result if we used `use_val_labs()`, but 
`uvl()` is more compact for typing and printing purposes.

### Non-standard Evaluation using `with_val_labs()` and `wvn`
labelr also offers an option to overlay ("swap out") value labels using 
`base::with()`-like non-standard evaluation. This is helpful in a few specific 
cases. 

```{r}
with(df, table(gender, raceth)) # base::with()

with_val_labs(df, table(gender, raceth)) # labelr::with_val_labs()

wvl(df, table(gender, raceth)) # labelr::wvl is a more compact alias
```

In a little bit, we'll see that we have some parallel options for overlaying 
("turning on") NAME labels.

### Add value labels back to the data.frame with `add_lab_cols()`
If all this wrapping and interactive toggling back and forth is making you 
dizzy, we could do something more permanent. 

For example, we can assign the result of a `use_val_labs()` call to an object. 
The result will be a data.frame with the same names and dimensions as the one 
supplied, with value labels replacing values for all value-labeled variables 
(or for a subset of those variables, if you specify them). Those variables will 
be coerced to character (if they were not already). Since there is no "undo" 
shortcut for this action, it is safest to assign the result to a new object.

```{r}
df_labd <- use_val_labs(df)
head(df_labd) # note, this is utils::head(), not labelr::headl()
```

Better still, we do not need to choose between values and labels. We can use 
`add_lab_cols()` to preserve all existing variables (columns), including the 
value-labeled ones, while adding to our data.frame an additional 
labels-as-values column for each value-labeled column. 

Easier done than said, perhaps. Take a look:

```{r}
df_plus_labs <- add_lab_cols(df)
head(df_plus_labs[c("gender", "gender_lab", "raceth", "raceth_lab")])
```

### "Filter values using labels" with `flab()` 
We also can filter a value-labeled data.frame using value labels, returning a 
subsetted data.frame in terms of the original values. In other words, we can use 
the more semantically meaningful value labels to guide our subsetting, even as 
they remain "invisible" and "in the background" as far as conventional R data 
views, descriptives, and analyses are concerned. Again, I find this "easier done 
than said."

```{r}
head(df)

df1 <- flab(df, raceth == "Asian" & gender == "Female")

head(df1, 5) # returned df1 is in terms of values, just like df

headl(df1, 5) # note use of labelr::headl; labels are there
```

We've used these two variables' value labels to guide our filtering, without 
ever explicitly changing the contents of our columns from values to labels. For 
instance, note that we did NOT make an explicit call to `use_val_labs()` or 
`add_lab_cols()` before our call to `flab()`. So long as we are providing 
actually existing value labels that have been previously applied to the columns 
in question, `flab()` knows where to find them and how to use them.

### "Subset using labels" with `slab()` 
As with `base::subset()`, we can also limit which columns we return. In this 
case, we filter on two value-labeled columns and return a data.frame consisting 
of only those columns.

```{r}
df2 <- slab(df, raceth == "Black" & gender == "Male", gender, raceth)
head(df2, 10)
```

In the case of `slab()`, we simply list the desired columns -- unquoted and 
comma-separated -- after the filter

## "Using" NAME labels 
Just as we used `use_val_labs()` to swap out values for value labels, we can 
use `use_name_labs()` to swap out variable names for variable NAME labels. Let's
illustrate this with the mtcars data.frame.

First we'll construct a vector of named labels.

```{r}
names_labs_vec <- c(
  "mpg" = "Miles/(US) gallon",
  "cyl" = "Number of cylinders",
  "disp" = "Displacement (cu.in.)",
  "hp" = "Gross horsepower",
  "drat" = "Rear axle ratio",
  "wt" = "Weight (1000 lbs)",
  "qsec" = "1/4 mile time",
  "vs" = "Engine (0 = V-shaped, 1 = straight)",
  "am" = "Transmission (0 = automatic, 1 = manual)",
  "gear" = "Number of forward gears",
  "carb" = "Number of carburetors"
)
```

Now, we will apply them to mtcars and assign the resulting data.frame to a new 
data.frame called mt2.

```{r}
mt2 <- add_name_labs(mtcars,
  vars = names(names_labs_vec),
  labs = names_labs_vec
)
```

Here is an alternative `add_name_labs()` syntax that would get us to the same 
end state:

```{r}
mt2 <- add_name_labs(mtcars,
  name.labs = c(
    "mpg" = "Miles/(US) gallon",
    "cyl" = "Number of cylinders",
    "disp" = "Displacement (cu.in.)",
    "hp" = "Gross horsepower",
    "drat" = "Rear axle ratio",
    "wt" = "Weight (1000 lbs)",
    "qsec" = "1/4 mile time",
    "vs" = "Engine (0 = V-shaped, 1 = straight)",
    "am" = "Transmission (0 = automatic, 1 = manual)",
    "gear" = "Number of forward gears",
    "carb" = "Number of carburetors"
  )
)
```

Now, let's swap out names for NAME labels.
```{r}
mt2 <- use_name_labs(mt2)

head(mt2[c(1, 2)])
```

Yikes, the longer column names stretch things out quite a bit. Even so, if we 
wish to keep our name labels "on" and work with them as our new column names, 
one approach is to use `get_name_labs` to get a look-up table, then use 
copy-and-paste or RStudio auto-complete capabilities to "hand jam" these into 
subsequent calls. 

For example: 
  
  ```{r}
lm(`Miles/(US) gallon` ~ `Number of cylinders`, data = mt2) # pasting in var names
lm(mpg ~ cyl, data = use_var_names(mt2)) # same result if name labels are "off"
```

While this works, freehand typing or copy-and-paste is clunky and quickly 
becomes tedious. There are other less painful ways we can use these NAME labels, 
once we've swapped them in for our original column names using `use_name_labs()` 
(as in the above example). For instance, we can take advantage of commands that 
work over all columns of a data.frame and, hence, don't require us to type 
individual column names. Here are a few illustrative examples.

```{r}
sapply(mt2, median) # get the median for every name-labeled variable

collapse::qsu(mt2) # use an external package for more informative descriptives
```

Another approach is to use `with_name_labs()` (or its more compact alias 
`wnl()`), which will automatically display name labels in place of column names 
in fairly flexible ways. `with_name_labs()` is an alternative approach that you 
can call on the regular, name-labeled data.frame. You should **not** call it on 
a data.frame after swapping in name labels with `use_name_labs()`.

With that said, let's revert back to our original column names, then we'll 
verify that the name labels are still there in the background, **then** we'll 
take `with_name_labs()` for a spin.  

```{r}
# invert our prior use_name_labs() call
mt2 <- use_var_names(mt2) # revert from name labels back to original colnames
head(mt2[c(1, 2)])
```

```{r}
# first, show that mt2 now has original column names swapped back in
head(mt2)

# verify that the name labels are still present and available in the background
get_name_labs(mt2)
```

Note that this sort of switching back and forth between your original column 
names and name labels (i.e., `use_name_labs()` and `use_var_names()`) assumes 
you are **not** otherwise modifying your either set of names in the interim. 

Now, pay attention to the variable names in the console output of the following 
calls to `with_name_labs()`.You'll be using the familiar column names in your 
function call expressions, but their corresponding name labels will appear in 
the console output.

```{r}
# demo with_name_labs() (note that with_name_labs() will achieve same result)
with_name_labs(mt2, t.test(mpg ~ am)) # wnl() is alias for with_name_labs()

with_name_labs(mt2, lm(mpg ~ am))

wnl(mt2, summary(mt2)) # wnl() is alias for with_name_labs()

wnl(mt2, xtabs(~gear)) # wnl() is alias for with_name_labs()

with(mt2, xtabs(~gear)) # compare this base::with() call to wnl() call above
```

Keep in mind that `with_name_labs()` is intended for self-contained calls 
involving exploratory analysis activities -- things like simple plots, 
descriptives, and models. The underlying function is based on simple regular 
expressions and **will throw an error** if you attempt to use it in contexts 
involving (1) exotic or non-standard operators, (2) multi-step workflows (e.g., 
pipes), OR (3) data management and cleaning commands. Still, as shown above, it 
plays well with a range of "workhorse" exploratory and descriptive commands.

## Alias Functions and Conclusion
This concludes our whirlwind tour of labelr functionalities. You've graduated.

Well, almost. Before you go, here is a list of aliases for common functions. 
Aside from having a different name, each alias function is identical to (i.e., 
performs the same operations, returning the same result as) the parent function 
that it aliases. More concise and more cryptic, these alias functions will save 
you some typing at the console (and some characters in your scripts).

The available aliases are as follows:
  
* `add_val_labs` alias is `avl`

* `get_val_labs` alias is `gvl` 

* `drop_val_labs` alias is `dvl` 

* `add_val1` alias is `avl1` 

* `drop_val1` alias is `dvl1` 

* `add_quant_labs` alias is `aql` 

* `add_quant1` alias is `aq1` 

* `add_m1_lab` alias is `am1l` 

* `use_val_labs` alias is `uvl`

* `with_val_labs` alias is `wvl`

* `add_lab_cols` alias is `alc`

* `add_lab_col1` alias is `alc1`

* `add_lab_dummies` is  `ald` 

* `add_lab_dumm1` is  `ald1` 

* `lab_int_to_factor` is  `int2f` 

* `factor_to_lab_int` is  `f2int` 

* `add_name_labs` is  `anl` 

* `get_name_labs` alias is `gnl`

* `drop_name_labs` alias is `dnl` 

* `use_name_labs` alias is `unl` 

* `use_var_names` alias is `uvn` 

* `with_name_labs` alias is `wnl`

* `with_both_labs` alias is `wbl`

* `add_frame_lab` alias is `afl`

* `get_frame_lab` alias is `gfl`

* `drop_frame_lab` alias is `dfl`

* `axis_lab` is  `alb` 
